---
title: "Tabernanthalog (TBG) Audit & Visual Makeover"
subtitle: "A Reproducible Analysis of David E. Olson Lab's 2021 Nature
      Publication"
author: "Isaac McGinn"
institution: "University of California Davis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    highlight-style: github
---
## 1. Introduction & Biological Question

Serotonergic psychedelics such as psilocybin, LSD, and **ibogaine** have proven to be profoundly effective in treating neuropsychiatric disorders, including substance use disorders. However, their therapeutic use is limited by their intense hallucinogenic side effects and, in the case of ibogaine, the risk of fatal cardiac arrhythmias through inhibition of the hERG potassium channel.

In order to overcome these barriers, the Olson Lab at the University of California, Davis developed **tabernanthalog (TBG)**, a water-soluble, non-hallucinogenic, and non-toxic analogue of ibogaine. TBG retains the neural plasticity properties of ibogaine while eliminating hallucinogenic effects and cardiac toxicity (Cameron et al., 2021).
      
In this computational reanalysis and visual redesign, I utilize raw, unpolished behavioral data deposited on Figshare (ID: 11634795) from the seminal publication:
      
 > Cameron, L.P., Tombari, R.J., Lu, J. *et al.* A non-hallucinogenic
      psychedelic analogue with therapeutic potential. *Nature* **589**,
      474–479 (2021).  [https://doi.org/10.1038/s41586-020-3008-z](https://doi.org/10.1038/s4158
  6-020-3008-z)
      
My primary objective is to audit Figure 4e of this study, verifying whether a single administration of TBG (10 mg/kg) significantly reduces daily alcohol consumption in mice over a multi-day binge-drinking timeline, and to replace the original "dynamite plots" with a modern, high-density visualization that reveals individual biological trajectories.

## Raw Data Import
```{r}
#load packages
library(tidyverse)
library(pzfx)
library(here)
#import raw data using 'here()' to find project root
raw_data <- read_pzfx(
  here("data/raw/extracted/Final Data for Publication/Fig 4e_Alcohol over time copy.pzfx"),
  table = "Data 1"
)
#print raw dim to verify import
dim(raw_data)
```
## Wrangle Data
```{r}
#preview raw_data 
glimpse(raw_data)
#tidy based on observations
tidy_data <- raw_data |> 
  pivot_longer(
    cols = -ROWTITLE,
    names_to = "mice_replicates",
    values_to = "alcohol_intake"
  ) |> 
  rename(day = ROWTITLE) |> 
  separate_wider_delim(
    cols = mice_replicates,
    delim = "_",
    names = c("treatment", "mouse_id")
  ) |> 
  mutate(
    treatment = as.factor(treatment),
    mouse_id = as.integer(mouse_id)
  )
#check to see that changes were made
print(tidy_data)
```
## Replication of Fig.e: Alcohol Intake Over Time 
```{r}
#group and summarize data 
summary_data <- tidy_data |> 
  group_by(day, treatment) |> 
  summarise(
    mean_intake = mean(alcohol_intake),
    sd_intake = sd(alcohol_intake),
    n = n(),
    sem_intake = sd_intake / sqrt(n)
  ) |> 
  ungroup()
#check changes 
print(summary_data)
#- changes made
#replicate figure using summary_data 
replication_plot <- ggplot(
  summary_data,
  aes(
      x = as.factor(day),
      y = mean_intake,
      group = treatment,
      fill = treatment,
      color = treatment
  )
) +
  #treatment lines
  geom_line(
    position = position_dodge(.2),
    linewidth = 1 
  ) +
  #SEM error bars
  geom_errorbar(
    aes(
      ymin = mean_intake - sem_intake,
      ymax = mean_intake + sem_intake
    ),
    width = .1,
    position = position_dodge(.2),
    linewidth = .8
  ) + 
  #filled points
  geom_point(
    shape = 21,
    size = 4,
    color = "black",
    stroke = 1,
    position = position_dodge(.2)
  ) +
  #color mapping 
  scale_fill_manual(values = c("VEH" = "white",
                               "TBG" = "steelblue")) +
  scale_color_manual(values = c("VEH" = "black",
                                "TBG" = "steelblue")) +
  #zooming without cut off error bars 
  coord_cartesian(
    ylim = c(0, 15)) +
  scale_y_continuous(
    breaks = seq(0, 15, 5)) +
  #labels 
  labs(
    x = "Days",
    y = "Alcohol Intake (g/kg/day)",
    title = "Recreation of Figure 4e"
  ) +
  theme_classic()
#display plot 
print(replication_plot)
```



## Statistical Audit 
```{r}
#fit 2-way ANOVA test 
anova_model <- aov(
  alcohol_intake ~ treatment * as.factor(day) +
Error(as.factor(mouse_id)),
  data = tidy_data 
  )
#print model
summary(anova_model)
```


